---
title: "NICE Framework Alignment Matrix"
subtitle: "Competency Mapping for Curriculum Approval"
description: "Detailed alignment between activities and NIST SP 800-181 Rev 1 (NICE Framework) competencies"
format:
  html:
    toc: true
    toc-depth: 3
  pdf:
    toc: true
  docx:
    toc: true
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(knitr)
library(kableExtra)
library(DT)

# Activity-to-Work-Role mapping
activity_roles <- tibble::tribble(
  ~Activity, ~`Primary Focus`, ~`Related Work Roles`, ~`Key Competency Areas`,
  "Security Detective Teams", "Threat detection & analysis", "Cyber Defense Analyst, Vulnerability Assessment Analyst", "Log analysis, pattern recognition, anomaly identification",
  "Ethics in Automated Security", "Policy & governance", "Cyber Policy Planner, Privacy Officer", "Policy development, stakeholder analysis, risk assessment",
  "AI-Assisted Incident Response", "Incident handling", "Incident Responder, Cyber Defense Analyst", "Incident coordination, response procedures, communication"
)

# Tasks data
tasks <- tibble::tribble(
  ~Code, ~Statement, ~Connection, ~Activity,
  "T1084", "Identify anomalous network activity", "Students analyze system logs to identify unusual patterns that may indicate security incidents", 1,
  "T1118", "Identify vulnerabilities", "Students discover weaknesses in security practices (e.g., password reuse patterns)", 1,
  "T1119", "Recommend vulnerability remediation strategies", "Students propose solutions after identifying security gaps", 1,
  "T1307", "Develop cybersecurity policy recommendations", "Students create policies balancing security with privacy/fairness", 2,
  "T1308", "Coordinate cybersecurity policy review and approval processes", "Students consider multiple stakeholder perspectives in policy design", 2,
  "T1605", "Advise management, staff, and users on cybersecurity policy", "Students present and defend policy recommendations", 2,
  "T1300", "Report cybersecurity incidents", "Students document and communicate incident findings", 3,
  "T1310", "Implement protective or corrective measures when a cybersecurity incident or vulnerability is discovered", "Students make decisions about containment and remediation", 3,
  "T1221", "Disseminate incident and other Computer Network Defense (CND) information", "Students coordinate information sharing across team roles", 3
)

# Knowledge data
knowledge <- tibble::tribble(
  ~Code, ~Statement, ~Connection, ~Activity,
  "K0682", "Knowledge of cybersecurity threats", "Students learn to recognize indicators of potential threats in log data", 1,
  "K0683", "Knowledge of cybersecurity vulnerabilities", "Students identify common vulnerability patterns (credential exposure, access anomalies)", 1,
  "K0684", "Knowledge of cybersecurity threat characteristics", "Students distinguish between normal and suspicious activity patterns", 1,
  "K0751", "Knowledge of system threats", "Students understand how attackers exploit system weaknesses", 1,
  "K0752", "Knowledge of system vulnerabilities", "Students recognize technical and human factors that create vulnerabilities", 1,
  "K0659", "Knowledge of information privacy technologies", "Students learn how AI systems handle personal data", 2,
  "K0736", "Knowledge of information technology (IT) security principles and practices", "Students apply security principles to policy decisions", 2,
  "K0682", "Knowledge of cybersecurity threats", "Students understand threats that automated systems address", 2,
  "K0683", "Knowledge of cybersecurity vulnerabilities", "Students recognize how policies can create or prevent vulnerabilities", 2,
  "K0724", "Knowledge of incident response principles and practices", "Students learn structured approach to incident handling", 3,
  "K0725", "Knowledge of incident response tools and techniques", "Students use AI as an incident response tool", 3,
  "K0726", "Knowledge of incident handling tools and techniques", "Students apply containment and eradication concepts", 3,
  "K0682", "Knowledge of cybersecurity threats", "Students identify threat actors and attack vectors", 3,
  "K0684", "Knowledge of cybersecurity threat characteristics", "Students understand attacker tactics, techniques, and procedures", 3
)

# Skills data
skills <- tibble::tribble(
  ~Code, ~Statement, ~Connection, ~Activity,
  "S0544", "Skill in recognizing vulnerabilities", "Core activity: students identify security weaknesses in evidence documents", 1,
  "S0800", "Skill in analyzing organizational patterns and relationships", "Students correlate data across multiple sources to build complete picture", 1,
  "S0874", "Skill in performing network traffic analysis", "Students analyze access logs and activity patterns", 1,
  "S0540", "Skill in identifying network threats", "Students determine which anomalies represent genuine security concerns", 1,
  "S0878", "Skill in performing risk analysis", "Students weigh security benefits against privacy/fairness risks", 2,
  "S0800", "Skill in analyzing organizational patterns and relationships", "Students consider how policies affect different stakeholder groups", 2,
  "S0850", "Skill in performing cost/benefit analysis", "Students evaluate trade-offs in automated security decisions", 2,
  "S0878", "Skill in performing risk analysis", "Students assess incident severity and business impact", 3,
  "S0540", "Skill in identifying network threats", "Students determine scope and nature of security incidents", 3,
  "S0800", "Skill in analyzing organizational patterns and relationships", "Students coordinate team response across multiple roles", 3
)

# Grade-band progression
grade_bands <- tibble::tribble(
  ~`Grade Band`, ~`Competency Focus`, ~`Cognitive Level`, ~`Activity Adaptation`,
  "K-2", "Awareness", "Recognition", "Identify \"helpers\" vs. \"bad actors\" in simple scenarios",
  "3-5", "Understanding", "Comprehension", "Explain why patterns matter and what makes activity suspicious",
  "6-8", "Application", "Analysis", "Analyze evidence with AI partnership, correlate multiple data sources",
  "9-12", "Evaluation", "Synthesis", "Critique AI recommendations, design policies, lead incident response"
)

# Cross-cutting: Human-AI Collaboration
human_ai_collab <- tibble::tribble(
  ~Competency, ~`NICE Alignment`, ~`Development Across Activities`,
  "Critical evaluation of AI outputs", "S0544, S0800", "All activities require students to verify AI findings",
  "Complementary task allocation", "T1084, T1118, T1310", "Students learn which tasks benefit from AI vs. human analysis",
  "Contextual judgment", "K0682, K0684", "Students provide context that AI systems cannot access"
)

# Cross-cutting: Professional Communication
prof_comm <- tibble::tribble(
  ~Competency, ~`NICE Alignment`, ~`Development Across Activities`,
  "Technical documentation", "T1300, T1221", "Students document findings and decisions",
  "Stakeholder communication", "T1605, T1308", "Students present findings to different audiences",
  "Team coordination", "T1310, T1221", "Students work in roles with distinct responsibilities"
)

# Helper function for consistent table formatting
# type: "tks" for Task/Knowledge/Skill tables (narrow Code column)
#       "overview" for Quick Reference and similar (balanced columns)
#       "crosscut" for Human-AI Collaboration and Professional Communication
format_tks_table <- function(df, type = "tks") {
  if (knitr::is_latex_output()) {
    tbl <- df |>
      kable(format = "latex", booktabs = TRUE)

    if (type == "tks") {
      tbl |>
        kableExtra::column_spec(1, width = "1.2cm") |>
        kableExtra::column_spec(2, width = "5cm") |>
        kableExtra::column_spec(3, width = "7cm")
    } else if (type == "overview") {
      tbl |>
        kableExtra::column_spec(1, width = "3cm") |>
        kableExtra::column_spec(2, width = "2.5cm") |>
        kableExtra::column_spec(3, width = "4cm") |>
        kableExtra::column_spec(4, width = "4cm")
    } else if (type == "crosscut") {
      tbl |>
        kableExtra::column_spec(1, width = "4cm") |>
        kableExtra::column_spec(2, width = "3cm") |>
        kableExtra::column_spec(3, width = "6cm")
    } else if (type == "grade") {
      tbl |>
        kableExtra::column_spec(1, width = "1.5cm") |>
        kableExtra::column_spec(2, width = "2.5cm") |>
        kableExtra::column_spec(3, width = "2.5cm") |>
        kableExtra::column_spec(4, width = "6cm")
    } else {
      tbl
    }
  } else if (knitr::is_html_output()) {
    df |>
      datatable(
        options = list(
          pageLength = 10,
          dom = "ft",
          ordering = TRUE
        ),
        rownames = FALSE,
        class = "display compact"
      )
  } else {
    df |>
      kable()
  }
}
```

## Overview

This document provides a detailed mapping between our Human-AI Partnership activities and the NICE Framework (NIST SP 800-181 Rev 1, Version 2.0.0) Task, Knowledge, and Skill (TKS) statements. These alignments support curriculum approval processes for K-12 cybersecurity education programs and demonstrate how classroom activities develop authentic workforce competencies.

::: {.callout-note}
## For Curriculum Administrators
This matrix demonstrates standards alignment for state/district cybersecurity curriculum approval. All TKS codes reference the official NICE Framework available at [niccs.cisa.gov](https://niccs.cisa.gov/workforce-development/nice-framework).
:::

## Quick Reference: Activity-to-Work-Role Mapping

```{r}
#| label: activity-roles
#| echo: false

activity_roles |>
  format_tks_table(type = "overview")
```

## Activity 1: Security Detective Teams

### Competency Alignment

Students practice foundational skills aligned with **Cyber Defense Analysis** and **Vulnerability Assessment** work areas.

#### Tasks Practiced

```{r}
#| label: activity1-tasks
#| echo: false

tasks |>
  filter(Activity == 1) |>
  select(Code, `Task Statement` = Statement, `Activity Connection` = Connection) |>
  arrange(Code) |>
  format_tks_table()
```

#### Knowledge Developed

```{r}
#| label: activity1-knowledge
#| echo: false

knowledge |>
  filter(Activity == 1) |>
  select(Code, `Knowledge Statement` = Statement, `Activity Connection` = Connection) |>
  arrange(Code) |>
  format_tks_table()
```

#### Skills Practiced

```{r}
#| label: activity1-skills
#| echo: false

skills |>
  filter(Activity == 1) |>
  select(Code, `Skill Statement` = Statement, `Activity Connection` = Connection) |>
  arrange(Code) |>
  format_tks_table()
```

## Activity 2: Ethics in Automated Security

### Competency Alignment

Students practice foundational skills aligned with **Cybersecurity Policy** and **Privacy** work areas.

#### Tasks Practiced

```{r}
#| label: activity2-tasks
#| echo: false

tasks |>
  filter(Activity == 2) |>
  select(Code, `Task Statement` = Statement, `Activity Connection` = Connection) |>
  arrange(Code) |>
  format_tks_table()
```

#### Knowledge Developed

```{r}
#| label: activity2-knowledge
#| echo: false

knowledge |>
  filter(Activity == 2) |>
  select(Code, `Knowledge Statement` = Statement, `Activity Connection` = Connection) |>
  arrange(Code) |>
  format_tks_table()
```

#### Skills Practiced

```{r}
#| label: activity2-skills
#| echo: false

skills |>
  filter(Activity == 2) |>
  select(Code, `Skill Statement` = Statement, `Activity Connection` = Connection) |>
  arrange(Code) |>
  format_tks_table()
```

## Activity 3: AI-Assisted Incident Response

### Competency Alignment

Students practice foundational skills aligned with **Incident Response** and **Cyber Defense** work areas.

#### Tasks Practiced

```{r}
#| label: activity3-tasks
#| echo: false

tasks |>
  filter(Activity == 3) |>
  select(Code, `Task Statement` = Statement, `Activity Connection` = Connection) |>
  arrange(Code) |>
  format_tks_table()
```

#### Knowledge Developed

```{r}
#| label: activity3-knowledge
#| echo: false

knowledge |>
  filter(Activity == 3) |>
  select(Code, `Knowledge Statement` = Statement, `Activity Connection` = Connection) |>
  arrange(Code) |>
  format_tks_table()
```

#### Skills Practiced

```{r}
#| label: activity3-skills
#| echo: false

skills |>
  filter(Activity == 3) |>
  select(Code, `Skill Statement` = Statement, `Activity Connection` = Connection) |>
  arrange(Code) |>
  format_tks_table()
```

## Grade-Band Progression

Each activity introduces NICE Framework competencies at developmentally appropriate levels, building complexity as students advance through grade bands:

```{r}
#| label: grade-bands
#| echo: false

grade_bands |>
  format_tks_table(type = "grade")
```

## Cross-Cutting Competencies

All three activities develop these foundational competencies:

### Human-AI Collaboration Skills

```{r}
#| label: human-ai-collab
#| echo: false

human_ai_collab |>
  format_tks_table(type = "crosscut")
```

### Professional Communication

```{r}
#| label: prof-comm
#| echo: false

prof_comm |>
  format_tks_table(type = "crosscut")
```

## Using This Matrix

### For Curriculum Approval

When preparing curriculum approval documentation, reference specific TKS codes to demonstrate standards alignment. Map activity learning objectives to corresponding NICE competencies, and document how grade-band progressions build systematically toward workforce readiness.

### For Assessment Design

Use TKS statements to create competency-based rubric criteria that align assessment evidence with specific skill demonstrations. This approach enables educators to track student progression across competency areas over time.

### For Career Pathway Planning

Connect classroom activities to authentic workforce competencies by introducing students to cybersecurity career language. These connections build awareness of diverse cybersecurity roles and help students envision potential career pathways.

## References

- NIST Special Publication 800-181 Revision 1: Workforce Framework for Cybersecurity (NICE Framework)
- NICE Framework Version 2.0.0 (September 2024)
- Available at: [https://niccs.cisa.gov/workforce-development/nice-framework](https://niccs.cisa.gov/workforce-development/nice-framework)
